Longitudinal Principal Manifold Estimation

### 2023-08-09 paper/main.tex
update literature review and augmentation explanation

Some previous work has been conducted on longitudinal dimension reduction methods, but this research has primarily focused on linear methods, such as the longitudinal principal component analysis approach proposed by \cite{kinsonLongitudinalPrincipalComponent2020}. To our knowledge, the approach to longitudinal smoothing in the manifold learning setting taken in this article is novel.

@@

Notably, the manifolds estimated in the simulated datasets included several self-intersecting manifolds, which methods based on the principal curve framework, including the principal manifold approach used here, are typically unable to fit due to the self-consistency property discussed above. To avoid this issue, we augmented the simulated datasets with polar or spherical coordinates, depending on the value of $D$. As a simple example, the unit circle can be considered a self-intersecting manifold with $d = 1$ when viewed in two dimensions. However, by adding a third dimension equal to the angle of each observation from the origin, we the manifold can be considered as having $d = 1$ and $D = 3$, while avoiding any self-intersections. This enables the PME algorithm, and thus the LPME algorithm, to be fit under these circumstances. This process is illustrated graphically in Figure \ref{fig:unit_circle_augmentation}.

\begin{figure}%
  \centering
  \subfloat[\centering $D = 2$]{{\includegraphics[width=9cm]{unit_circle_D2}}}%
  \hfill
  \subfloat[\centering $D = 3$]{{\includegraphics[width=9cm]{unit_circle_D3}}}
  \caption{Augmenting the Unit Circle}
  \label{fig:unit_circle_augmentation}
\end{figure}

\zielinski{The simulation results shown below use partial results run on my computer.}

@@

\begin{table}
  \centering
  \begin{tabular}{|c c c c c c c c c c|}
    \hline
    Method & Case 1 & Case 2 & Case 3 & Case 4 & Case 5 & Case 6 & Case 7 & Case 8 & Case 9 \\
    \hline
    Data & 0.222 (0.261) & 0.513 (0.411) & N/A & 27.3 (74.9) & 0.986 (0.787) & 1.51 (6.04) & N/A & N/A & N/A \\
    LPME & 0.129 (0.161) & 0.417 (0.960) & N/A & 19.0 (63.0) & 0.776 (0.777) & 1.27 (5.71) & N/A & N/A & N/A \\
    PME & 0.247 (0.435) & 0.854 (1.62) & N/A & 27.3 (74.9) & 1.04 (1.07) & 1.55 (6.04) & N/A & N/A & N/A \\
    PC/PS & 0.188 (0.247) & 0.600 (0.303) & N/A & 27.3 (74.9) & 0.939 (0.730) & 1.04 (2.20) & N/A & N/A & N/A \\
    \hline
  \end{tabular}
  \caption{MSD Comparison to True Values, Mean (SD).}
  \label{table:simulation_results_mean}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{|c c c c c c c c c c|}
    \hline
    Method & Case 1 & Case 2 & Case 3 & Case 4 & Case 5 & Case 6 & Case 7 & Case 8 & Case 9 \\
    \hline
    Data & 0.144 (0.222) & 0.466 (0.658) & N/A & 4.17 (19.4) & 0.890 (1.39) & 0.305 (1.07) & N/A & N/A & N/A \\
    LPME & 0.0727 (0.129) & 0.250 (0.477) & N/A & 3.23 (11.3) & 0.575 (1.08) & 0.293 (0.907) & N/A & N/A & N/A \\
    PME & 0.123 (0.249) & 0.516 (0.757) & N/A & 4.13 (19.2) & 0.897 (1.39) & 0.331 (1.04) & N/A & N/A & N/A \\
    PC/PS & 0.112 (0.205) & 0.556 (0.427) & N/A & 4.12 (19.3) & 0.805 (1.26) & 0.567 (0.388) & N/A & N/A & N/A \\
    \hline
  \end{tabular}
  \caption{MSD Comparison to True Values, Median (IQR).}
  \label{table:simulation_results_median}
\end{table}

@@

While the LPME algorithm demonstrated strong performance in a number of settings, it also has several limitations. First, because the method explicitly builds on the results of an estimate provided by the PME algorithm, the LPME algorithm will only fit properly in situations where PME fits properly. Because the output of PME is highly dependent on the success of its initial parameterization, in practice this limits the success of LPME to situations where Isomap performs well, which excludes nonconvex manifolds, manifolds that have holes, or manifolds that are too curvy. The option to use alternative methods, such as Locally Linear Embedding or Laplacian Eigenmaps, to generate the initial parameterization for the PME algorithm may help to alleviate this concern. There is also room to improve the tuning of the initialization method, which may result in better performance, especially when considering data with different levels of noise present.

>>>>

Some previous work has been conducted on longitudinal dimension reduction methods, but this research has primarily focused on linear methods, such as the longitudinal principal component analysis approach proposed by \cite{kinsonLongitudinalPrincipalComponent2020}. \cite{wolzManifoldLearningBiomarker2010a} used Laplacian eigenmaps to to incorporate longitudinal information when reducing neuroimaging information to lower dimensional spaces, but estimated parameterizations for baseline images and longitudinal image differences separately, where we seek to parameterize them simultaneously. \cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling. However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability. To our knowledge, the approach to longitudinal smoothing in the general-purpose manifold learning setting taken in this article is novel.

@@

Notably, the manifolds estimated in the simulated datasets included several self-intersecting manifolds, which methods based on the principal curve framework, including the principal manifold approach used here, are typically unable to fit due to the self-consistency property discussed above. To avoid this issue, we augmented the simulated datasets with polar or spherical coordinates, depending on the value of $D$. As a simple example, the unit circle can be considered a self-intersecting manifold with $d = 1$ when viewed in two dimensions. However, by adding a third dimension equal to the angle of each observation from the origin, the manifold can be considered as having $d = 1$ and $D = 3$, while avoiding any self-intersections. This enables the PME algorithm, and thus the LPME algorithm, to be fit under these circumstances. This process is illustrated graphically in Figure \ref{fig:unit_circle_augmentation}.

This approach to data augmentation offers an alternative to the method used in \cite{mengPrincipalManifoldEstimation2021} to fit the PME algorithm to self-intersecting manifolds, where the full dataset was manually partitioned to avoid intersections within each partition. Reaching a combined estimate after fitting PME to each partition individually requires "gluing" the estimated manifolds on each partition together, which may introduce error to the combined manifold estimate. This source of error is avoided using the data augmentation approach. Additionally, scaling the added dimensions by a constant value enables the modification of the level of curvature present in the manifold. This may encourage improved performance when estimating manifolds that have too much curvature for PME to reach an appropriate fit under normal circumstances. However, an in depth comparison of these approaches is beyond the scope of this article and is left for future work.

\begin{figure}%
  \centering
  \subfloat[\centering Unaugmented Unit Circle]{{\includegraphics[width=9cm]{unit_circle_D2}}}%
  \hfill
  \subfloat[\centering Augmented Unit Circle]{{\includegraphics[width=9cm]{unit_circle_D3}}}
  \caption{Augmenting the Unit Circle}
  \label{fig:unit_circle_augmentation}
\end{figure}

@@

\begin{table}
  \centering
  \begin{tabular}{|c c c c c c c c c c|}
    \hline
    Method & Case 1 & Case 2 & Case 3 & Case 4 & Case 5 & Case 6 & Case 7 & Case 8 & Case 9 \\
    \hline
    Data & 0.222 (0.255) & 0.512 (0.409) & N/A & 30.5 (89.0) & 0.975 (0.770) & 1.43 (6.13) & N/A & N/A & N/A \\
    LPME & 0.125 (0.161) & 0.384 (0.659) & N/A & 27.8 (266) & 0.788 (0.843) & 1.18 (5.71) & N/A & N/A & N/A \\
    PME & 0.267 (0.862) & 0.839 (1.93) & N/A & 30.5 (89.2) & 1.03 (1.07) & 1.47 (6.14) & N/A & N/A & N/A \\
    PC/PS & 0.188 (0.244) & 0.599 (0.297) & N/A & 30.4 (89.0) & 0.930 (0.713) & 1.00 (2.14) & N/A & N/A & N/A \\
    \hline
  \end{tabular}
  \caption{MSD Comparison to True Values, Mean (SD).}
  \label{table:simulation_results_mean}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{|c c c c c c c c c c|}
    \hline
    Method & Case 1 & Case 2 & Case 3 & Case 4 & Case 5 & Case 6 & Case 7 & Case 8 & Case 9 \\
    \hline
    Data & 0.145 (0.231) & 0.462 (0.661) & N/A & 4.21 (20.7) & 0.888 (1.33) & 0.279 (1.06) & N/A & N/A & N/A \\
    LPME & 0.0733 (0.122) & 0.245 (0.460) & N/A & 3.27 (12.8) & 0.579 (1.08) & 0.269 (0.890) & N/A & N/A & N/A \\
    PME & 0.129 (0.258) & 0.508 (0.747) & N/A & 4.19 (20.7) & 0.881 (1.36) & 0.313 (1.04) & N/A & N/A & N/A \\
    PC/PS & 0.117 (0.205) & 0.560 (0.416) & N/A & 4.18 (20.7) & 0.817 (1.21) & 0.556 (0.386) & N/A & N/A & N/A \\
    \hline
  \end{tabular}
  \caption{MSD Comparison to True Values, Median (IQR).}
  \label{table:simulation_results_median}
\end{table}

@@

While the LPME algorithm demonstrated strong performance in a number of settings, it also has several limitations. First, because the method explicitly builds on the results of an estimate provided by the PME algorithm, the LPME algorithm will only fit properly in situations where PME fits properly. Because the output of PME is highly dependent on the success of its initial parameterization, in practice this limits the success of LPME to situations where Isomap performs well, which excludes nonconvex manifolds, manifolds that have holes, or manifolds that have too high a level of curvature. The option to use alternative methods, such as Locally Linear Embedding or Laplacian Eigenmaps, to generate the initial parameterization for the PME algorithm may help to alleviate this concern. There is also room to improve the tuning of the initialization method, which may result in better performance, especially when considering data with different levels of noise present.
